---
title: "ST340CW1Mihir"
author: "u1729346"
date: "22/10/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Question 1a: Merge Implementation

```{r merge}
mergeLR <- function(left, right){
  ret <- numeric(length(left) + length(right))
  i <- 1
  j <- 1
  k <- 1
  while(i <= length(left) || j <= length((right))) {
    if(i <= length(left) && j <= length((right))){
      if(left[i] < right[j]){
        ret[k] <- left[i]
        i <- i + 1
        k <- k + 1
      }
      else{
        ret[k] <- right[j]
        j <- j + 1
        k <- k + 1
      }
    }
    else if(i <= length(left)){
      ret[k] <- left[i]
      k <- k + 1
      i <- i + 1 
    }
    else if(j <= length(right)){
      ret[k] <- right[j]
      k <- k + 1
      j <- j + 1     
    }
  }
  return(ret)
}
```
\textbf{Merge Test}:
```{r testm}
left <- c(1,2,2.34)
right <- c(-12,-5,9,2000)
print(mergeLR(left,right))
```

### Question 1b: Mergesort Implementation

```{r mergesort}
mergesort <- function(a){
  if(length(a) == 1){
    return(a)
  }
  else{
    mid <- floor(length(a)/2)
    left <- numeric(mid)
    right <- numeric(length(a) - mid)
    for(i in 1:length(left)){
      left[i] <- a[i]
      }
    for(j in 1:length(right)){
      right[j] <- a[j + mid]
    }
    return(mergeLR(mergesort(left), mergesort(right)))
  }
}
```
\textbf{MergeSort Test}:
```{r testms}
input <- c(1,2,2.34,2000,-12,-5,9)
print(mergesort(input))
```

### Question 1d: Mergesort Correctness

\textbf{Proposition}: Mergesort correctly sorts an array ($a[1],...,a[n]$) of size n.

\textbf{Base Case}: When $n=1$, $a[1]$ is sorted by default since it contains only one element. 

\textbf{Inductive Hypothesis}: When $n=k$, the array is correctly sorted by mergesort. i.e. $a[1]\leq ... \leq a[k]$.

\textbf{Inductive Step}: When $n=k+1$,
$$mid=\lfloor{\frac{k+1}{2}}\rfloor$$
$$left = a[1],...,a[mid]$$
$$right = a[mid+1],...,a[k+1]$$

Since $\lfloor{\frac{k+1}{2}}\rfloor \leq k$ and $k+1-(\lfloor{\frac{k+1}{2}}\rfloor + 1) = k - \lfloor{\frac{k+1}{2}}\rfloor<k$, left and right will be correctly sorted by mergesort as assumed in our inductive hypothesis. We must show that mergeLR correctly outputs a sorted array when called upon left and right.

\textbf{Proposition}: mergeLR correctly returns a single array with the elements of the two input arrays in sorted order. Suppose left and right are two arrays with lengths p and q respectively and the algorithm is on its $k^{th}$ iteration.

\textbf{Base Case}: When $k =1$, 
$$ret = concat(min\{left[1],right[1]\}, max\{left[1],right[1]\})$$
Hence, the returned array is in correct sorted order.

\textbf{Inductive Hypothesis}: For arbitrary $k$, the returned merged array is correctly sorted. i.e. $ret[1] \leq ...\leq ret[k]$. 

\textbf{Inductive Step}: For the $(k+1)^{th}$ iteration, suppose i is an index running through left, s.t. $1\leq i\leq p$ and j is an index running through right, s.t. $1\leq j\leq q$. Assume $left[i]\leq right[j]$ such that $ret[k] = left[i]$ and using the inductive hypothesis, we have $ret[1] \leq ...\leq ret[k]$. Considering the assumed case, 
$$ret[k+1] = min\{left[i+1], right[j]\}$$
$$\implies ret[1]\leq ...\leq ret[k]\leq ret[k+1]$$
since $left[i]\leq left[i+1]\leq right[j]$ and in the case where $i+1 > p$, the if condition fails and the else if condition $j\leq length(right)$ is activated, where
$$ret[k+1] = right[j],...,ret[p+q]=right[q]$$
and since right was already sorted, we have
$$ret[k+1]\leq ...\leq ret[p+q]$$
and hence the merged array is correctly sorted.

For the case when $right[j] < left[i]$, the proof is identical due to the symmetry of the problem.

### Question 1d: Mergesort Runtime 

Given an array of size n, mergesort is recursively called upon two equal halves (since array is assumed even) and once each half is of size 1, merge is executed $O(n)$ times, until the original array size is attained. Hence, we arrive at the recurrence equation given by:
$$T(1) = 1,$$
$$T(n) = 2T(\frac{n}{2}) + O(n),\;\;\; n > 1$$
\textbf{Proposition}: $T(n) \leq n\log_{2}(n), \;\; \forall n\in\{2^{k}:k\in \mathbb{N}\}$

\textbf{Base Case}: $n = 2^{1}$
$$T(2) = 2T(1) + O(2) = 2 \leq 2\log_{2}(2) = 2$$
Hence, the base case is true.
 
\textbf{Inductive Hypothesis}: Assume true for $n = 2^{k}$:
$$T(2^{k}) = 2T(\frac{2^{k}}{2}) + O(2^{k}) $$
$$\;\; = 2T(2^{k-1}) + O(2^{k}) \leq 2^{k}log_{2}(2^{k})$$

\textbf{Inductive Step}: For $n = 2^{k+1}$:
$$T(2^{k+1}) = 2T(\frac{2^{k+1}}{2}) + O(2^{k+1})$$
$$\;\; = 2T(2^{k}) + O(2^{k+1})$$
using our inductive hypothesis, we have:
$$\;\; \leq 2^{k+1}\log_{2}(2^{k+1}))$$
and hence the proposition stands $\forall n \in \{2^{k}: k \in \mathbb{N}\}$.

### Question 1e: Quicksort vs Mergesort

Quicksort and Mergesort are both divide and conquer based sorting strategies. Quicksort uses a pivot (either chosen naively or by other methods such as median) to partition the array whereas Mergesort always partitions the array into halves. This results in Mergesort having a worst-case time complexity of $O(n\log(n))$. A scenario where the array is sorted in descending order results in Quicksort's worst-case time complexity of $O(n^{2})$. Hence, Mergesort is faster and more efficient than Quicksort.

## Question 2
## Majority element  

```{r majorityElement}

majElement <- function(myData)
{
  if (length(myData) == 1)
  {
    return(myData)
  }
  left <- myData[1:floor(length(myData)/2)]
  right <- myData[(floor(length(myData)/2)+1):length(myData)]
  majLeft <- majElement(left)#Recursive call 1
  majRight <- majElement(right)#Recursive call 2
  count<-0
  if(majLeft != "No Majority")
  {
    count <- length(myData[myData==majLeft])
    for(each in 1:count) #First n equivalence relations
    {
        if(each == (length(myData)/2)+1)
        {
          return(majLeft)
        }
    }
  }
  count <- 0
  if(majRight != "No Majority")
  {
    count <- length(myData[myData==majRight])
    for(each in 1:count) #Second n equivalence relations
    {
       if(each == (length(myData)/2)+1)
       {
         return(majRight)
       }
    }
  }
  return("No Majority")
}
```

### 2A:
The program works by splitting the input list into two sublists, then calling itself recursively on each of the two sublists. If the provided list is of length 1, it returns the element, which must be the majority element. The principle of the solution is that if a majority element exists then it must be the majority element of at least one of the two sublists. Likewise, if neither sublist has a majority element then the superlist cannot have a majority element*. The returned values from the recursive call form a set of at most 2 potential majority elements. We may then count the elements to identify the majority element. Because the problem statement does not allow the use of ">" or "<", identifying if the count of a given element is sufficent to make it the majority requires checking if each natural number up to the count is equal to the required number of elements. 

*Proof: Suppose the list has length $n$. Then the majority element must have at least $\frac{n}{2} + 1$ instances. If the statement does not hold then there must exist some a and b such that a + b =  $\frac{n}{2} + 1$ and a $\leq$ b < $\frac{n}{4}$ + $\frac{1}{2}$. However the right equation can be rearranged to a + b < $\frac{n}{2} + 1$, violating the left equation. Therefore the statement must hold.

### 2B:
We have that the algorithm will split a given input list into two lists, each half the size, and call itself again. We also have that there will be 2n equivalence relations per call of the function, where n is the length of the input list. Since we have $n \in \{2^k : k \in N \}$ we may prove an upper bound on the number of equivalence relations of the algorithm using induction on k.

Thus we have that T(n) is equal to $2n + 2*T(\frac{n}{2})$. Proof $2n\log_{2}n$ is an upper bound:

Proof for k = 1:

$T(2^k) = T(2) = 2*2 + 2*(0) = 4 \leq 2(2)\log_{2}2$

Inductive step:

Suppose the statement is true for k = x. For k = x+1:

$$\begin{aligned} T(2^{x+1}) &= 2*2^{x+1} + 2*T(x)\\
            &= 2*2^{x+1} + 2*(2*2^{x}\log_{2}2^{x})\\
            &= 2*2^{x+1}(1 + \log_{2}2^{x})\\
            &= 2*2^{x+1}(1 + \log_{2}2^{x+1} - 1)\\
            &= 2*2^{x+1}(\log_{2}2^{x+1})\\
            \square 
            \end{aligned}$$

Therefore $2n\log_{2}n$ is an upper bound for all $n \in \{2^k : k \in N \}$.
